Race-based disparities in allocation of academic disciplinary actions are associated with county-level rates of bias

### 2018-05-18 written_docs/draft.tex
.gitignore is now working

@@

We used three distinct data sources for the work described here. The
academic disciplinary data is part of the Civil Rights Data Collection
(CRDC) through the US Department of Education. The dataset we used comes
from the 2013-2014 academic year and has data on \enquote{all public
local and educational agencies and schools, including long-term secure
juvenile justice facilities, charter schools, alternative schools, and
schools serving students with disabilities.} In total, the CRDC data
represents 95507 institutions enrolling approximately 50 million
students, of which approximately 25 million are white and 7.8 million
are Black\footnote{We note that there are a number of differences
  between the analyses we registered and those presented in the main
  text. Our general conclusions are largely the same for both sets of
  analyses. We opted to report the modified analyses for reasons of
  clarity and to remain congruent with previous research on the same
  topics. The registered analyses can be found, in full, in the
  appendix.}. Previous work using these data have identified a number of
districts whose data are in error, and have excluded juvenile justice
facilities, as these institutions constitute dramatically different
educational environments, where the meaning of disciplinary actions may
be quite different (Losen et al., 2015). After these exclusions are
applied, the final sample used for modeling consists of 90002
institutions, enrolling 32 million black or white students, of which 25
million are white and 7 million are black.

We obtained county-level demographic information for use as covariates
and state-level demographic information for use in post-stratification
from the US Census Bureau. For both county-level and state-level
demographics, we use 5-year estimates for the period ending in 2014 from
the American Community Survey, which surveys around 295k households per
month.

We used measurements of implicit and explicit bias available from data
collected through Project Implicit (Xu, Nosek, \& Greenwald, 2014). From
this data, we used only respondents who had geographic information that
would allow us to place them in a United States county, identified as
White, and visited the site before 2015. This consisted of approximately
1.1 million total respondents from 3091 counties.

A subset of years in the Project Implicit data also collected
occupational information from respondents. As identified in our
pre-analysis plan, we took advantage of the presence of primary and
secondary educators in these data to test whether any associations
between bias and race-based differences in the rates of disciplinary
action were stronger among these respondents. Filtering for only white
individuals who identified as primary, secondary, special education, and
other teachers and instructors (occupation codes 25-2000 and 25-3000)
reduced the dataset to 63552 respondents. In order to assure that our
estimates were reasonably stable, we limited analysis to only counties
that had at least 50 respondents. As such, our teacher analysis is
limited to just 287 counties. Additionally, because we do not know of
any state-level demographic estimates for teachers, we are unable to
perform post-stratification for these data.

\subsection{Measures}\label{measures}

The primary outcome in this analysis is a count of the number of
students by race (black and white) who received one of several types of
disciplinary action. We report here rates of out-of-school suspension,
in-school suspension, school-related arrests, law enforcement referrals,
and total number of expulsions of any type.

For both the post-stratification procedure (described below) and the
actual statistical models used for inference, we used the same set of
covariates. These covariates were at the state-level for
post-stratification and were at the county-level for the statistical
models used for final estimation and inference. Specifically, our
population based covariates were the total population count, the
proportion of the population that is black, the proportion of the
population that is white and the ratio of black-to-white persons. We
also used socioeconomic covariates. We used the percent of individuals
aged 16 or over who were in the labor force but unemployed, the median
household income, and the percentage of all families whose income is
below the poverty line.

Finally, for the implicit and explicit bias measures, we relied on two
primary variables from the Project Implicit data. Implicit bias was
assessed via an Implicit Association Test (IAT). This test uses a
speeded dual-categorization task in which individuals must quickly
categorize black and white faces and \enquote{good} and \enquote{bad}
words with key presses. The difference in how quickly and accurately
participants are able to pair white faces with \enquote{good} words and
black faces with \enquote{bad} words in comparison to the inverse is
thought to reflect implicit associations between the two races and
positive and negative affective reactions. This association is indicated
in the IAT D-score, which we used as a measure of implicit bias. Our
measure of explicit bias is the difference between reported warmth
towards whites (i.e. \emph{how warm or cold do you feel towards Whites?
0=very cold, 10=very warm}) and reported warmth toward blacks.

@@

Data analysis proceded in two steps. We first estimated county-level
implict and explicit bias using multilevel regression and
post-stratification. Post-stratification is a valuable procedure in
obtaining accurate geographical population-based estimates because it
allows a non-representative sample (e.g Project Implicit) to more
closely resemble the true population, and it regularizes extreme
observations with little data to support them (e.g.~a county with only a
handful of respondents with especially high or low scores) (Gelman \&
Little, 1997; Park, Gelman, \& Bafumi, 2004). Following past work
(Leitner, Hehman, Ayduk, \& Mendoza-Denton, 2016), we identified age as
one dimension along which IAT respondents differed from the general
population in ways that could bias our conclusions (Gonsalkorale,
Sherman, \& Klauer, 2009). Our post-stratification weighting scheme is
as follows: We first grouped respondents into five age group categories
(15-24, 25-34, 35-54, 55-75, and 75+). We next fit multilevel models
estimating bias (implicit and explicit biases seperately) as a function
of our state-level covariates (the \enquote{fixed} effects), and allowed
the estimates to vary by age bin, county, and state (the
\enquote{random} effects). Next, we determined the population of whites
in each county in these age groups using the American Community Survey's
5-year estimates ending in 2014. Finally, we used our estimated models
to predict the expected response for each age bin, in each county. Our
final county-level estimates are the average of the values predicted for
the 5 age bins, weighted by the population size of that bin in that
county. As a result of this procedure, we can be confident that our
county-level estimates should more closely approximate what our
estimates would look like if the Project Implicit data were truly
representative along the age dimension in all counties.

After obtaining these estimates, we use them as predictors in bayesian
multilevel logistic regressions. Formally, the likelihood for a given
observation is written as a binomial function:

\[
{n\choose y}\pi^y(1-\pi)^{n-y}
\]

Where \(y\) is the observed count of incidents (e.g.~number of black or
white students suspended), \(n\) is the number of at-risk students
(e.g.~total number of black or white students), and
\(\pi = g^{-1}(\eta)\) is the probability of the incident ocurring. For
this analysis, the linear predictor takes the form of a multilevel model
with a set of effects that vary over county:

\[
\eta = \alpha + X\beta + \gamma_{county}
\]

Where \(\alpha\) is an intercept that is constant across observations,
\(\beta\) represents a set of effects that are also constant across
observations (i.e.~fixed effects), and \(\gamma_{county}\) represents
intercepts and effects of ethnicity that vary across the counties
(i.e.~random efffects).

In addition to the covariates described above, we also include effects
of race, implicit bias, explicit bias, and the two-way interactions
between implicit bias and race and explicit bias and race. We fit
separate models for each of the outcomes.

@@

All numerical predictor variables were standardized at the appropriate
level (county, state) before model estimation to help with estimation
efficiency and interpretability. We set priors for the intercept and
coefficients in the bayesian models to be weakly informative normal
distributions centered on zero with a standard deviation of five. All
other parameters were left to default values. Data analysis was done in
R (R Core Team, 2016) version 3.3.2 running under OS X 10.11.6.
Post-stratification was done with lme4, version 1.1.14 (Bates, Mächler,
Bolker, \& Walker, 2015). Final model fitting was done on the university
cluster running Springdale Linux, release 6.9 using rstanarm, version
2.17.2 (Stan Development Team, 2016). We used the implementation of the
consensus monte carlo algorithm found in parallelMCMCcombine, version
1.0 (Miroshnikov \& Conlon, 2014). Figures were made with ggplot2,
version 2.2.1 (Wickham, 2009), with data manipulation done using dplyr
version 0.7.2 (Wickham, Francois, Henry, \& Müller, 2017) and tidyr,
version 0.7.1 (Wickham \& Henry, 2017). A full report of session
information can be found on the OSF page (\ldots{}.)

@@

We first report the results of estimating the implicit and explicit
biases in from Project Implicit data. Overall, the individuals in
project implicit show a pro-white bias in both implicit (\emph{mean} =
0.40, \emph{sd} = 0.41), and explicit measures (\emph{mean} = 0.88
\emph{sd} = 1.83). When aggregated at the county level and adjusted with
poststratification, the summary statistics across counties are similar
in terms of their location, but as expected, the variability is much
diminished (\(mean_{implicit}\) = 0.40, \(sd_{implicit}\) = 0.02;
\(mean_{explicit}\) = 0.79, \(sd_{explicit}\) = 0.15, where on both
scales 0 = no bias, and positive numbers indicate a pro-white bias).

@@

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:disc-count}Count of students by race receiving each type of disciplinary action}
\begin{tabular}{lll}
\toprule
group & \multicolumn{1}{c}{metric} & \multicolumn{1}{c}{students}\\
\midrule
black & expulsions & 36,755.00\\
black & school arrests & 21,456.00\\
black & in-school suspension & 829,706.00\\
black & law enforcement referral & 53,306.00\\
black & out-of-school suspension & 1,850,492.00\\
white & expulsions & 55,832.00\\
white & school arrests & 21,420.00\\
white & in-school suspension & 1,054,172.00\\
white & law enforcement referral & 81,565.00\\
white & out-of-school suspension & 1,768,326.00\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\end{table}

Table \ref{tab:disc-count} shows the number of students of each race who
were reported having received each of the actions under consideration.
The counts range from a low of just 21420 white students arrested to a
high of 1850492 black students receiving an out-of-school suspension.
Considering the vast differences in the overall number of black and
white students, this simple count already illustrates that black
students are disciplined at rates far higher than their white
counterparts.

\subsection{Associations across
county}\label{associations-across-county}

Figure \ref{fig:overall-associations} shows the estimate of primary
interest for each of the models. The estimates displayed are the
coefficients for the interaction between race and each of the two bias
measurements. Given that African Americans are the baseline group,
negative values for this coefficient indicate that as the bias in a
county increases, the gap between the probability of a black student
being disciplined and the probability of a white student being
disciplined grows.

@@

Several patterns are apparent from this figure. First, with the
exceptions of implicit bias and expulsions and law enforcement
referrals, all estimates are directionally consistent with higher levels
of bias leading to larger differences between groups. Second, these
effects are especially consistent for the two types of suspensions. The
largest effect estimated is between implicit bias and out-of-school
suspensions. The difference in the slope of the association between
implicit bias and the log of the odds for out-of-school suspensions
between white and black students is estimated to be -0.25, with 95\% of
the posterior distribution between -0.31 and -0.20 and a proportion
\textgreater{}.99 of the posterior distribution consistent with a
negative effect.

Although not nearly as large of a difference, we have similar certainty
with respect to the association between out-of-school suspensions and
explicit bias. The difference in the slope of the association between
explicit bias and the log of the odds for out-of-school suspensions
between white and black students is estimated to be -0.08, with 95\% of
the posterior distribution between -0.14 and -0.03 and a proportion
\textgreater{}.99 of the posterior distribution consistent with a
negative effect.

The estimated associations for in-school suspensions are smaller still,
but are generally consistent with effects of the same direction. For
implicit bias, the relevant parameter is estimated at -0.02 {[}-0.06,
0.01{]} \(p_{neg}\) = 0.93, and for explicit bias, the effect is
slightly larger, and a positive effect is essentially not credible,
given the data and model -0.05 {[}-0.08, -0.02{]} \(p_{neg}\)
\textgreater{}.99.

Other outcomes are estimated with less precision, or with patterns that
are inconsistent between implicit and explicit bias. For instance,
examining the associations for expulsions, the effect of explicit biases
are generally in the expected direction (\emph{est} = -0.07, {[}-0.15,
0.02{]}, \(p_{neg}\) = .94), but the effect for implicit bias is
estimated to be close to zero, with a enough uncertainty (\emph{est} =
0.04, {[}-0.05, 0.12{]}, \(p_{neg}\) = .19) to make it difficult to
claim an effect of one direction or the other. The model indicates
similar uncertainty with respect to school-related arrests and both
estimates of bias (\(est_{explicit}\) = -0.02, {[}-0.11, 0.07{]},
\(p_{neg}\) = .67; \(est_{implicit}\) = -0.02, {[}-0.12, 0.08{]},
\(p_{neg}\) = .66).

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/detail-fig-exp-1.pdf}
\caption{\label{fig:detail-fig-exp}Association between bias and expulsions.
Top: Association between bias and the estimated probability of
expulsion. Line is the mean of the posterior. Bands indicate 95\%
uncertainty intervals; Bottom: Association between bias and the relative
risk ratio for black students to white students. Points represent
counties, whose size are scaled to the number of schools in that
county.}
\end{figure}

To better illustrate the nature of these relationships, figure
\ref{fig:detail-fig-exp} shows the estimated probabilities of expulsion
for black and white students as a function of bias, along with the
relative risk ratio for black students. The relative risk ratio is the
ratio of the probability that a black student will be expelled to the
probability that a white student will get expelled. Values over 1
reflect higher levels of punishment for black students. As previously
indicated in table \ref{tab:disc-count}, the top part of this figure
makes plain the higher probability of expulsion for black children. In a
county at the mean of the distribution of bias, approximately 0.15\% of
black students are expected to be expelled {[}0.14, 0.17{]}. The
corresponding rate for white students is much lower, with about 0.06\%
expected to be expelled {[}0.06, 0.07{]}. Moving to a county one
standard deviation above the mean of explicit bias has the effect of
increasing the estimated percentage of black students expected to be
expelled to 0.16\% {[}0.13, 0.18{]}, while the percentage of white
students expected to be expelled would decline to 0.06\% {[}0.05,
0.07{]}. The same movement for implicit bias would slightly increase the
expected expulsions for black students ( 0.15\% {[}0.13, 0.18), and
increase the expected expulsions for white students a very small amount
more to (0.06\% {[}0.06, 0.07). In real terms, in a county at the mean
of the distributions of bias, for every white student expelled, we
should expect 2.45 {[}2.25, 2.45{]} black students to be expelled. If we
move to a county one standard deviation above the mean of explicit bias,
the ratio of black to white students expelled increases to 2.63 {[}2.33,
2.63{]}, while the same movement for implicit bias slightly decreases
the ratio to 2.36 {[}2.13, 2.36{]}.

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/detail-fig-susp-1.pdf}
\caption{\label{fig:detail-fig-susp}Association between bias and
out-of-school suspensions Top: Association between bias and the
estimated probability of suspension Line is the mean of the posterior.
Bands indicate 95\% uncertainty intervals; Bottom: Association between
bias and the relative risk ratio for black students to white students.
Points represent counties, whose size are scaled to the number of
schools in that county.}
\end{figure}

>>>>

In comparison to White Americans, African Americans exhibit poorer
educational outcomes across a range of metrics. One outcome of
particular concern is the gap in disciplinary actions (Kinsler, 2011;
Okonofua, Walton, \& Eberhardt, 2016). Research using administrative
datasets and longitudinal samples clearly show that African American
students are far more likely to be suspended or expelled (Aud,
KewalRamani, \& Frohlich, 2011; Yeager, Purdie-Vaughns, Hooper, \&
Cohen, 2017), and conditional on an office referral, are more likely to
receive stiffer punishments (J. F. Gregory, 1995). Concerns about these
disparities are exacerbated by their associations with long-term life
outcomes, including prospects for employment (Pager, Western, \& Sugie,
2009), mental and physical health (Pascoe \& Smart Richman, 2009), and
involvement in the criminal justice system (Hirschfield, 2009). In this
paper, we aim to establish that these disciplinary gaps are related to
environmental levels of bias in predictable ways.

It is perhaps not surprising that levels of bias would be associated
with racial disciplinary gaps. This straightforward statement, however,
is contextualized in a number of ways. First, the operationalization of
bias should play a role in the association. As we describe below, levels
of implicit and explicit biases should be differentially associated with
disciplinary actions. This differential association should be
determined, in part, by the type of disciplinary action under
examination. For a given student committing a specific offense, the
educational institution may opt for one of several disciplinary actions.
These actions vary in severity, institutional costs, and the degree of
formal administrative involvement.

To illustrate this problem in the context of previous research, we now
describe research on the associations between implicit bias and explicit
bias and decision making. This sets the stage for understanding what
types of decisions administrators and educators make when electing a
specific disciplinary action. We then describe the types of disciplinary
actions that are taken and the characteristics of these actions in terms
of severity, institutional costs, and formal administrative involvement.
Finally, in crossing this information, we concretely specify the set of
associations expected when examining bias and disciplinary actions in
American schools.

\subsection{Bias and decision-making}\label{bias-and-decision-making}

Previous research on biased attitudes and behavior has shown that
although implicit and explicit attitudes are typically correlated
(Hofmann, Gawronski, Gschwendner, Le, \& Schmitt, 2005; McConnell \&
Leibold, 2001), they tend to behaviorally manifest slightly differently.
As a default, a naiive expectation is that explicit attitudes should
guide behavior, since if people tell us they like something, we should
expect that to generally correspond to their behavior. However, under
some circumstances, implicit biases are more strongly associated with
behavioral outcomes. B. A. Nosek, Hawkins, and Frazier (2011) suggest
that, in comparison to explicit attitudes, implicit attitudes are more
strongly associated with behavior when the individual lacks motivated to
explicitly respond differently than their implicit attitude would
suggest, when they do not have the opportunity or ability to modify
their initial response inclination, or when they do not have awareness
of the implicit response, the behavior, or the link between them. In
contrast, the more motivated, more able, or more aware the person is
over their behavioral response, the more explicit biases should play a
role. One key situational factor that determines the response is
situational uncertainty. In uncertain situations, the respondent is not
likely to have strong motivations to change their initial response,
otherwise the situation would not be uncertain. Research supports this
idea, as when the situation is characterized by greater uncertainty,
implicit biases are known to be stronger determinants of behavior
(Friese, Hofmann, \& Schmitt, 2009; Galdi, Arcuri, \& Gawronski, 2008;
Gawronski, Geschke, \& Banse, 2003). In short, when taking (or not
taking) an action can be attributed to multiple reasons due to
situational ambiguity, then implicit biases should result in higher
rates of discriminatory behavior.

Accordingly, this research suggests that disciplinary actions that are
implemented quickly, have relatively little administrative oversight,
and require little in the way of insitutional resources should be
characterized by strong associations with implicit bias. On the other
hand, any disciplinary action that requires higher levels of oversight,
resources, or lengthy time investments to complete should be more
strongly associated with explicit bias.

\subsection{Types of disciplinary
actions}\label{types-of-disciplinary-actions}

We report here on five different types of disciplinary actions:
in-school suspensions, out-of-school suspensions, law enforcement
referrals, school-related arrests, and expulsions. As described above,
these actions vary in severity, required resources, and administrative
oversight.

The most extreme outcome, expulsion, is defined as when a student is
prohibited from ever returning to the educational institution. The
institution may or may not set up alternative educational services for
the student. Because of the the severe nature of this punishment, it
also requires considerable administrative oversight, and is often
conferred for behaviors that are relatively unambiguous (e.g.~bringing a
gun to school). We believe similar cases can be made for law enforcement
referrals and school-related arrests insofar as these actions tend to be
more severe, and require resources, or occasionally the involvement of
outside agencies.

Somewhat less severe than expulsions, but one of the most frequently
utilized disciplinary actions, out-of-school suspensions occupy
something of a middle ground. There is a substantial body of research
documenting the negative associations between this particular
disciplinary action and life outcome (Cuellar \& Markowitz, 2015;
Matjasko, 2011), as well as their disproportionate use on stigmatized
students (Haight, Gibson, Kayama, Marshall, \& Wilson, 2014; Kinsler,
2011). Out-of-school suspensions are administered for a wide range of
behaviors. Some of these behaviors are apt to be determined with a fair
degree of subjectivity (e.g.~defiance) (A. Gregory \& Weinstein, 2008),
while others may be more unambiguous (e.g.~fighting, drug use). In
comparison to expulsions and arrests, this type of action demands less
administrative oversight.

In-school suspensions, wherein the student is removed from the classroom
but remains under the direct supervision of school personnel, are also
administered for a wide range of behaviors. However, in-school
suspensions have been highlighted as a promising alternative to more
exclusionary types of disciplinary actions (e.g.~out-of-school
suspensions, expulsions) partly because students remain under the
supervision of school personnel, which keeps them in a structured,
educational setting (Anyon et al., 2014). Like out-of-school
suspensions, this type of action requires relatively little
administrative oversight. But unlike out-of-school suspension, removing
a student from a classroom to a \emph{different} educational setting
requires additional space and personnel, and so this represents an
inherent investment in the student.

\subsection{Concrete predictions of
association}\label{concrete-predictions-of-association}

Given these disciplinary outcomes and what is known about implicit and
explicit attitudes, what should we expect in examining associations
between them? First, it's clear that we should expect African American
students to receive higher levels of disciplinary actions. Second, we
should expect the disparity in disciplinary actions to generally be
positively associated with explicit bias. Finally, we should expect
implicit bias to only show associations in instances where there is
little administrative overhead, and where the behaviors to-be-punished
are often ambiguous. By this logic, out-of-school suspensions are the
only metric that meet these criteria.

\subsection{The present study}\label{the-present-study}

The present study uses a national dataset of disciplinary outcomes to
evaluate the degree to which implicit and explicit bias are associated
with each of these outcomes. After an initial, registered analytic
approach, we subsequently made several adjustments to strengthen our
analyses and streamlined them with respect to previous research. In
general, the data support the conclusion that implicit biases are
specifically associated with out-of-school suspensions, while explicit
biases tend to be associated with disciplinary actions of all types.

@@

\subsubsection{Disciplinary actions}\label{disciplinary-actions}

To assess rates of disciplinary action, we used data from the Civil
Rights Data Collection (CRDC) conducted by the US Department of
Education. The dataset we used comes from the 2013-2014 academic year
and has data on \enquote{all public local and educational agencies and
schools, including long-term secure juvenile justice facilities, charter
schools, alternative schools, and schools serving students with
disabilities.} In total, the CRDC data represents 95507 institutions
enrolling approximately 50 million students, of which approximately 25
million are white and 7.8 million are Black\footnote{We note that there
  are a number of differences between the analyses we registered and
  those presented in the main text. Our general conclusions are largely
  the same for both sets of analyses. We opted to report the modified
  analyses for reasons of clarity and to remain congruent with previous
  research on the same topics. The registered analyses can be found, in
  full, in the appendix.}. Previous work using these data have
identified a number of districts whose data are in error, and have
excluded juvenile justice facilities, as these institutions constitute
dramatically different educational environments, where the meaning of
disciplinary actions may be quite different (Losen et al., 2015). After
these exclusions are applied, the final sample used for modeling
consists of 90002 institutions, enrolling 32 million black or white
students, of which 24.70 million are white and 7.30 million are black.
From these data, we focus on the number of students by race (black and
white) who received one of several types of disciplinary action. We
report here rates of out-of-school suspension, in-school suspension,
school-related arrests, law enforcement referrals, and total number of
expulsions of any type.

\subsubsection{Racial Bias}\label{racial-bias}

We used measurements of implicit and explicit bias available from data
collected through Project Implicit (Xu, Nosek, \& Greenwald, 2014). For
a full description of the implicit and explicit bias measures available
in these data, refer to Xu et al. (2014) and Leitner, Hehman, Ayduk, and
Mendoza-Denton (2016). We note only that we used the IAT D-score as a
measure of implicit bias, and the difference between reported warmth
towards whites and warmth towards black (both measured from 0=very cold,
10=very warm) as a measure of explicit bias. Additionally, we used only
respondents who had geographic information that would allow us to place
them in a United States county, identified as White, and visited the
site before 2015. This consisted of approximately 1.1 million total
respondents from 3091 counties.

\subsubsection{Covariates}\label{covariates}

Each county-level variable used as a covariates in the final model and
the corresponding state-level variable used as a predictor in the
post-stratification scheme were taken from the from the same source.
Population size and proportions, socioeconomic indicators, and mobility
were all taken from the American Community Survey (ACS) 5-year estimates
for the time period ending in 2014. Urban-rural indicators were taken
from the 2010 US Census, and crime rates were taken from the FBI Uniform
Crime Reporting program, as made available through the National Archive
of Criminal Justice Data for each year from 2010-2014. Each of these
variables is described below.

\paragraph{Population size and
proportions}\label{population-size-and-proportions}

We obtained the total population, the proportion of the population that
is white, the proportion of the population that is black, and the ratio
of black-to-white people in the population from ACS table B02001.

\paragraph{Socioeconomic indicators}\label{socioeconomic-indicators}

We obtained estimates for the percentage of population with a Bachelor's
degree or higher, the percentage of the population aged 16 or over in
the labor force that is unemployed, the median household income, and the
percentage of families and people whose income in the last year was
below the poverty line from the ACS table DP03.

\paragraph{Urban-rural indicator}\label{urban-rural-indicator}

We obtained estimates of housing density per square mile of land area
from Census table GCT-PH1.

\paragraph{Mobility}\label{mobility}

We obtained estimates of population mobility by summing the percentage
of African Americans who moved from a different county, state, or
country into the county of interest (county-level covariate) or who
moved from a different state or country into the state of interest
(state-level covariate). We took these metrics from the ACS table S0701.

\paragraph{Crime}\label{crime}

We computed estimates of the number of violent crimes per person by
taking the number of crimes reported divided by the population size for
each year, and averaging the resulting proportions across the 5 years of
data.

@@

Because Project Implicit is a nonrandom sample, we used multilevel
regression and post-stratification to obtain accurate geographical
population-based estimates of implicit and explicit bias. Multilevel
regression and post-stratification corrects for biased sampling and
regularizes extreme observations with little data to support them
(e.g.~a county with only a handful of respondents with especially high
or low scores) (Gelman \& Little, 1997; Park, Gelman, \& Bafumi, 2004).
Following past work (Leitner et al., 2016), we identified age as one
dimension along which IAT respondents differed from the general
population in ways that could bias our conclusions (Gonsalkorale,
Sherman, \& Klauer, 2009). Our post-stratification weighting scheme is
as follows: We first grouped respondents into five age group categories
(15-24, 25-34, 35-54, 55-75, and 75+). We next fit multilevel models
estimating bias (implicit and explicit biases seperately) as a function
of our state-level covariates (the \enquote{fixed} effects: proportion
of population that is white; proportion of the population that is black,
the ratio of black to white people in the population, the percentage of
college graduates, the unemployment rate, the median household income,
the poverty rate, housing density, mobility rate, \& crime rate; each of
these is described in more detail below), and allowed the estimates to
vary by age bin, county, and state (the \enquote{random} effects). Next,
we determined the population of whites in each county in these age
groups using the American Community Survey's 5-year estimates ending in
2014. Finally, we used our estimated models to predict the expected
response for each age bin, in each county. Our final county-level
estimates are the average of the values predicted for the 5 age bins,
weighted by the population size of that bin in that county. As a result
of this procedure, we can be confident that our county-level estimates
should more closely approximate what our estimates would look like if
the Project Implicit data were truly representative along the age
dimension in all counties.

We analyzed these data using a series of bayesian multilevel logistic
regressions. We modeled the probability that a student would be expelled
as a function of a set of effects that are constant across observations
(i.e.~fixed effects: race (dummy coded), implicit bias, explicit bias,
an interaction term between race and implicit bias, an interaction term
between race and explicit bias, and all covariates described above) and
a set of effects that vary across counties (i.e.~random effects: overall
intercept \& race). We fit separate models for each of the outcomes. All
numerical predictors were standardized at the appropriate level (county,
state) before model estimation (for post-stratification as well as for
final inference) to help with estimation efficiency and
interpretability. We also set priors for the intercept and coefficients
in the bayesian model to be weakly informative normal distributions
centered on zero with a standard deviation of five. This corresponds to
a prior belief of all parameters taking values between -15 and 15.
Values outside this range are extremely unlikely given that all
variables were standardized prior to estimation. All other parameters
were left to default values.

@@

@@

We first report the results of estimating the implicit and explicit
biases in from Project Implicit data. When examining the unstandardized
county-level estimates adjusted with poststratification, we see that
there is a pro-white bias in both implicit (\(mean\) = 0.40, \(sd\) =
0.02) and explicit measures (\(mean\) = 0.82, \(sd\) = 0.17, where on
both scales 0 = no bias, and positive numbers indicate a pro-white
bias).

@@

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:disc-count}Percentage of students of each race receiving each type of disciplinary action}
\begin{tabular}{lll}
\toprule
metric & \multicolumn{1}{c}{black} & \multicolumn{1}{c}{white}\\
\midrule
expulsions & 0.5\% & 0.23\%\\
school arrests & 0.3\% & 0.09\%\\
in-school suspension & 11.47\% & 4.29\%\\
law enforcement referral & 0.76\% & 0.33\%\\
out-of-school suspension & 26.19\% & 7.2\%\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\end{table}

Table \ref{tab:disc-count} shows the percentage of students of each race
who were reported having received each of the actions under
consideration. Our statistical models (described below) show that these
differences are extremely unlikely to be due to chance.

\subsection{Associations across
counties}\label{associations-across-counties}

Figure \ref{fig:overall-associations} shows the estimate of primary
interest for each of the models. The estimates displayed are the
coefficients for the interaction between race and each of the two bias
measurements. Given that African Americans are the baseline group,
negative values for this coefficient indicate that as one moves into
counties with higher levels of bias, the gap between the probability of
a black student being disciplined and the probability of a white student
being disciplined grows. Table \ref{tab:reg-coefs} displays the
estimated coefficients and 95\% uncertainty intervals for each of the
fixed effects for each model.

@@

Several patterns are apparent from this figure. First, in all cases but
school-related arrests, county-level explicit biases are reliably
associated with disciplinary disparities, such that counties with more
bias are expected to have a larger racial gap. The effect with the
largest magnitude of these is out-of-school suspensions (\(M_{slope}\) =
-0.12, {[}-0.18, -0.06{]}), followed by expulsions (\(M_{slope}\) =
-0.11, {[}-0.21, -0.01{]}), law enforcement referrals (\(M_{slope}\) =
-0.10, {[}-0.16, -0.03{]}), and in-school suspensions (\(M_{slope}\) =
-0.08, {[}-0.12, -0.05{]}). The model indicates that the evidence for
school-related arrests is inconclusive (\(M_{slope}\) = -0.03, {[}-0.13,
0.08{]}).

Second, the effect with the largest magnitude of all is that between
implicit biases and out-of-school suspension (\(M_{slope}\) = -0.22,
{[}-0.28, -0.16{]}). All other estimated parameters are inconclusive.

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:reg-coefs}Regression coefficient estimates for the population-level (i.e. fixed) effects, along with 95\% uncertainty intervals for each of the disciplinary metrics.}
\small{
\begin{tabular}{llllll}
\toprule
 & \multicolumn{1}{c}{Out-of-School Suspension} & \multicolumn{1}{c}{Law Enf. Referral} & \multicolumn{1}{c}{In-School Suspension} & \multicolumn{1}{c}{School-Related Arrest} & \multicolumn{1}{c}{Expulsion}\\
\midrule
Intercept & -2.02, [-2.07,-1.97] & -5.75, [-5.83,-5.67] & -2.3, [-2.33,-2.26] & -7.82, [-7.98,-7.66] & -6.5, [-6.6,-6.4]\\
total population & 0.01, [-0.02,0.03] & 0.03, [-0.03,0.09] & -0.04, [-0.07,-0.01] & 0.13, [0.04,0.22] & 0.03, [-0.04,0.11]\\
propotion black & 0.4, [0.31,0.49] & -0.25, [-0.46,-0.03] & 0.19, [0.09,0.3] & -0.14, [-0.5,0.24] & -0.13, [-0.38,0.12]\\
proportion white & 0.03, [-0.03,0.09] & -0.2, [-0.36,-0.04] & 0, [-0.08,0.08] & 0.02, [-0.26,0.3] & -0.2, [-0.39,-0.01]\\
black-white ratio & -0.26, [-0.32,-0.19] & 0.07, [-0.1,0.24] & -0.34, [-0.42,-0.27] & -0.09, [-0.38,0.18] & -0.18, [-0.39,0.01]\\
college grads & -0.03, [-0.06,0] & 0.03, [-0.05,0.11] & -0.09, [-0.13,-0.05] & 0.17, [0.04,0.31] & -0.12, [-0.22,-0.02]\\
income & -0.13, [-0.18,-0.08] & -0.08, [-0.2,0.03] & -0.13, [-0.18,-0.07] & 0.03, [-0.15,0.21] & -0.28, [-0.42,-0.14]\\
poverty & -0.06, [-0.13,0.01] & -0.63, [-0.81,-0.46] & 0.1, [0.01,0.18] & -0.29, [-0.59,0] & -0.4, [-0.61,-0.19]\\
unemployment & 0.23, [0.19,0.27] & 0.2, [0.1,0.29] & 0.03, [-0.01,0.07] & 0.3, [0.15,0.46] & 0.29, [0.19,0.39]\\
crime & 0.04, [0.01,0.07] & 0.07, [0,0.14] & 0.06, [0.03,0.09] & 0.11, [-0.01,0.22] & 0.09, [0.01,0.17]\\
housing density & -0.01, [-0.03,0.02] & -0.07, [-0.14,0.01] & 0, [-0.04,0.03] & -0.06, [-0.15,0.04] & 0, [-0.07,0.08]\\
mobility & 0.02, [-0.01,0.05] & 0.07, [0,0.14] & 0.01, [-0.02,0.04] & 0.05, [-0.07,0.17] & 0.01, [-0.07,0.1]\\
race: white & -0.65, [-0.69,-0.61] & -0.87, [-0.93,-0.81] & -0.85, [-0.87,-0.82] & -1.1, [-1.22,-0.99] & -0.93, [-1.01,-0.85]\\
implicit bias & 0.36, [0.29,0.43] & -0.1, [-0.23,0.02] & 0.21, [0.15,0.28] & -0.25, [-0.45,-0.05] & 0.03, [-0.12,0.18]\\
explicit bias & -0.05, [-0.11,0.02] & -0.01, [-0.13,0.1] & 0.1, [0.04,0.16] & 0.37, [0.19,0.55] & 0.04, [-0.09,0.18]\\
implicit bias*race: white & -0.22, [-0.28,-0.16] & 0.05, [-0.02,0.12] & 0.01, [-0.02,0.05] & -0.01, [-0.13,0.09] & 0.07, [-0.03,0.17]\\
explicit bias*race: white & -0.12, [-0.18,-0.06] & -0.1, [-0.16,-0.03] & -0.08, [-0.12,-0.05] & -0.03, [-0.13,0.08] & -0.11, [-0.2,-0.02]\\
\bottomrule
\end{tabular}
}
\end{threeparttable}
\end{center}
\end{table}

The findings for other disciplinary metrics suggest a differential
effect of explicit and implicit bias. For example, there is a reliable
association between in-school suspensions and explicit bias. The
difference in the slope of the association between explicit bias and the
log of the odds for in-school suspensions between white and black
students is estimated to be -0.12, with 95\% of the posterior
distribution between -0.18 and -0.06 and a proportion \textgreater{}.99
of the posterior distribution consistent with a negative effect. Similar
effects can be seen for explicit bias and law enforcement referrals
(-0.10, {[}-0.16, -0.03{]}, \(p_{neg}\) \textgreater{}.99), and explicit
bias and expulsions (-0.11, {[}-0.21, -0.01{]}, \(p_{neg}\) = .99). In
contrast, zero is included as a plausible value for the associations
between implicit bias and in-school suspensions (0.01, {[}-0.02,
0.05{]}, \(p_{neg}\) = .27), law enforcement referrals (0.05, {[}-0.02,
0.12{]}, \(p_{neg}\) = .08), and expulsions (0.07, {[}-0.03, 0.17{]},
\(p_{neg}\) = .07).

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/detail-risks-1.pdf}
\caption{\label{fig:detail-risks}Association between bias and relative risk
ratio for black students to white students. Line is the mean of the
posterior. Bands indicate 95\% uncertainty intervals. Points represent
counties, whose sizes are scaled to the number of schools in that
county. Note that the y axis is cut at 8 for legibility despite data
extending beyond.}
\end{figure}
